Staged Abstract Interpreters

### 2019-03-26 paper/20_prelim.tex
paper writing

\paragraph{The Interpreter} Now we can define the semantics-agnostic interpreter
in monadic form. The essential idea is to traverse the abstract syntax tree
while maintaining the effects such as reader and state. 
It is worth noting that the interpreter is written in open-recursive style -- it
can not refer to itself directly, instead, @eval@ takes an additional parameter
@ev@ of type @Expr => Ans@ to refer to itself. Accordingly, the method @close@
that lifts lambda terms to closures and @ap_clo@ that applies functions also
takes an extra @ev@.

\begin{lstlisting}
  def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
    case Lit(i) => num(i)                      case Let(x, rhs, e) => for {
    case Var(x) => for {                         v <- ev(rhs)
      ρ <- ask_env                               α <- alloc(x)
      σ <- get_store                             _ <- set_store(α → v)
     } yield get(σ, ρ, x)                        rt <- ext_env(ev(e))(x → α)
    case Lam(x, e) => for {                    } yield rt
      ρ <- ask_env                             case Aop(op, e1, e2) => for {
    } yield close(ev)(Lam(x, e), ρ)              v1 <- ev(e1)                                               
    case App(e1, e2) => for {                    v2 <- ev(e2)
      v1 <- ev(e1)                             } yield arith(op, v1, v2)
      v2 <- ev(e2)                             case Rec(x, rhs, e) => for {
      rt <- ap_clo(ev)(v1, v2)                   α <- alloc(x)
    } yield rt                                   v <- ext_env(ev(rhs))(x → α)
    case If0(e1, e2, e3) => for {                _ <- set_store(α → v)
      cnd <- ev(e1)                              rt <- ext_env(ev(e))(x → α)
      rt <- br0(cnd, ev(e2), ev(e3))           } yield rt                    
    } yield rt
  }
\end{lstlisting}

>>>>

\paragraph{The Interpreter} Now we can define the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}.
The essential idea is to traverse the abstract syntax tree while maintaining the
effects such as reader and state.
It is worth noting that the interpreter is written in open-recursive style -- it
can not refer to itself directly, instead, @eval@ takes an additional parameter
@ev@ of type @Expr => Ans@ to refer to itself. Accordingly, the method @close@
that lifts lambda terms to closures and @ap_clo@ that applies functions also
takes an extra @ev@.

\vspace{-1em}
\begin{figure}[h!]
  \centering
  \begin{lstlisting}
            def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
              case Lit(i) => num(i)                    case Let(x, rhs, e) => for {
              case Var(x) => for {                       v <- ev(rhs)
                ρ <- ask_env                             α <- alloc(x)
                σ <- get_store                           _ <- set_store(α → v)
              } yield get(σ, ρ, x)                      rt <- ext_env(ev(e))(x → α)
              case Lam(x, e) => for {                  } yield rt
                ρ <- ask_env                           case Aop(op, e1, e2) => for {
              } yield close(ev)(Lam(x, e), ρ)            v1 <- ev(e1)                                               
              case App(e1, e2) => for {                  v2 <- ev(e2)
                v1 <- ev(e1)                           } yield arith(op, v1, v2)
                v2 <- ev(e2)                           case Rec(x, rhs, e) => for {
                rt <- ap_clo(ev)(v1, v2)                 α <- alloc(x)
              } yield rt                                 v <- ext_env(ev(rhs))(x → α)
              case If0(e1, e2, e3) => for {              _ <- set_store(α → v)
                cnd <- ev(e1)                            rt <- ext_env(ev(e))(x → α)
                rt <- br0(cnd, ev(e2), ev(e3))         } yield rt                    
              } yield rt
            }
  \end{lstlisting}
\caption{The generic interpreter,
  shared by the unstaged/staged + concrete/abstract interpreter.}
\label{fig:shared_int}
\end{figure}
\vspace{-1em}

### 2019-03-26 paper/30_stageconc.tex
paper writing

Readers may notice that the conversion between unstaged monads and monads is 
simply changing the unstaged data to staged data, which in fact can be achieved without
changing the implementation of monads. This is true so far, but as we will see
when introducing the nondeterminism monad (@ListT@) for abstract interpreters, as
the whole list will be a next-stage value instead of merely the elements in the list.
For this reason, we explicitly distinguish and present the two version of
monads, as a more consistent approach.

>>>>

Readers may notice that the conversion between unstaged monads and monads is 
simply changing the unstaged data to staged data, which in fact can be achieved without
changing the implementation of monads. This is true so far, but as we will see
it not that easy when introducing the nondeterminism monad (@ListT@) for
abstract interpreters, as the whole list will be a next-stage value instead of
merely the elements in the list.
For this reason, we explicitly distinguish and present the two version of
monads, as a more consistent approach.

### 2019-03-26 paper/40_absinterp.tex
paper writing

A complete lattice over set $S$ is a tuple of $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ declares the type class for lattices.
Similar to @MonadOps@ in Section \ref{generic_if}, we introduce an additional
higher-kinded type @R[_]@ to annotate the binding-times.

@@

In this section, we simply let @R[T] = T@ (i.e., @NoRep@) to instantiate the
unstaged lattice type class. An simple instance of lattice we used is the power
set abstract domain, where the ordering is subset relation, join is the union
operation, and meet is the intersect operation. The power set lattice can be
implemented as follows:

@@

Our instantiation roughly follows the Abstracting Abstract Machines methodology
\cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12} that transforms the
concrete semantics to abstract semantics. The store now maps addresses to sets
of abstract values, meaning that an address points to all possible values that
may occur at runtime, where the abstract value is either a closure or a single
abstract number \texttt{NumV}, which stands for the top element of the number
lattice. The address space is also constrained to be finite, in the monovariant
setting, we simply use variable names for address.

\begin{lstlisting}
trait Abstract extends Semantics {
  case class Addr(x: Ident);  sealed trait AbsValue
  case class CloV($\lambda$: Lam, $\rho$: Env) extends AbsValue
  case class NumV() extends AbsValue
  type Value = Set[AbsValue]
  type Env = Map[Ident, Addr];  type Store = Map[Addr, Value]
}
\end{lstlisting}

@@

\begin{lstlisting}
object AbsInterp extends Abstract {
  type R[+T] = T
  val $\rho$0 = Map[Ident, Addr]();  val $\sigma$0 = Map[Addr, Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = 
    $\sigma$.getOrElse(a, Lattice[Value].bot)
  def put($\sigma$: Store, a: Addr, v: Value): Store =
    $\sigma$ + (a -> (v $\sqcup$ get($\sigma$, a)))
  def alloc($\sigma$: Store, x: Ident): Addr = Addr(x)
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = Set(CloV($\lambda$, $\rho$))
  def num(i: Lit): Value = Set(NumV())
  def apply_closure(ev: EvalFun)
    (fs: Value, arg: Value, $\sigma$: Store): Ans = { var $\sigma$_* = $\sigma$
      val vs = for (CloV(Lam(x, e), $\rho$) <- fs) yield {
        val $\alpha$ = alloc($\sigma$_*, x)
        val (v, v$\sigma$) = ev(e, put($\rho$, x, $\alpha$), put($\sigma$_*, $\alpha$, arg))
        $\sigma$_* = v$\sigma$ $\sqcup$ $\sigma$_*; v
      }
    (vs.reduce(Lattice[Value].$\sqcup$), $\sigma$_*)
  }
  def branch0(cnd: Value, thn: =>Ans, els: =>Ans): Ans = 
    thn $\sqcup$ els
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = 
    Set(NumV())
} // to be continued
\end{lstlisting}

@@

@@

\begin{lstlisting}
case class CacheFix(evev: EvalFun => EvalFun) {
  var in = Map[Config, Ans](); var out = Map[Config, Ans]()
  def cached_ev(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = {
    val cfg: Config = (e, $\rho$, $\sigma$)
    if (out.contains(cfg)) out(cfg)
    else {
      val ans0 = in.getOrElse(cfg, Lattice[(Value, Store)].bot)
      out = out + (cfg -> ans0)
      val ans1 = evev(cached_ev)(e, $\rho$, $\sigma$)
      out = out + (cfg -> (ans0 $\sqcup$ ans1));  ans1
    }
  }
  def iter(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = {
    in = out; out = Map[Config, Ans](); cached_ev(e, $\rho$, $\sigma$)
    if (in == out) out((e, $\rho$, $\sigma$)) else iter(e, $\rho$, $\sigma$)
  }
}
override def eval_top(e: Expr, $\rho$: Env, $\sigma$: Store): Ans = 
  CacheFix(eval).iter(e, $\rho$, $\sigma$)
\end{lstlisting}

>>>>

A complete lattice over set $S$ is a tuple of $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ defines the type class of stage polymorphic
lattices. Similar to @MonadOps@ in Section \ref{generic_if}, we introduce an
additional higher-kinded type @R[_]@ to annotate the binding-times.

@@

In this section, we simply let @R[T] = T@ (i.e., @NoRep@) to instantiate the
unstaged lattice type class. An simple instance of lattice we used is the power
set abstract domain, where the ordering is subset relation, the join is union
operation, and the meet is intersect operation. The power set lattice can be
implemented as follows:

@@

% \cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12}

We follow the \textit{abstracting definitional interpreters} (ADIs) approach
\cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our concrete interpreter to
an abstract interpreter. We first widen the concrete values to @AbsValue@: numbers
are lifted to a singleton object @IntTop@ representing the set of all numbers;
closures remain the same. Then the type @Value@ is sets of @AbsValue@ (we abuse
the type names a little bit, as the generic interface defined the abstract type
@Value@).
The address space is constrained to be finited to ensure the computatbility of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After setteling the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@ respectively.
It is importatnt that the abstract store now maps addresses to sets of abstract
values, meaning an address may points to an over-approximated set of values
occurred at runtime.

\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ...
  }
\end{lstlisting}

A cache is a mapping from configurations to sets of value-store pairs, where the
configuration is a triple of current expression being evaluated, environment and store.

\begin{lstlisting}
  type Config = (Expr, Env, Store); type Cache = Map[Config, Set[(Value, Store)]]
  type Result = (R[List[(Value, Store)]], R[Cache])
\end{lstlisting}

@@

\begin{lstlisting}
\end{lstlisting}

@@

\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]
  def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def update_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

@@
